A fast algorithm for global code motion of congruent instructions

### 2017-11-11 cgo/paper.tex
codegen

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{graphviz}
\usepackage{auto-pst-pdf}
\usepackage{etoolbox}
\usepackage{flushend}
\usepackage{needspace}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage[final]{pdfpages}

@@

\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float tmin, tmax;
  float inv = 1.0f / d;
  if (inv >= 0) {
    tmin = (min - a) * inv;
    tmax = (max - a) * inv;
  } else {
    tmin = (max - a) * inv;
    tmax = (min - a) * inv;
  }
  return tmax + tmin;
}
\end{verbatim}

@@

Now if the computation of $tmax$ and $tmin$ are hoisted outside the
conditionals, the C code version would look like:
\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  float tmin = x, tmax = y;
  if (inv < 0) { tmin = y; tmax = x; }
  return tmax + tmin;
}
\end{verbatim}

@@

\section{A fast algorithm for code motion congruent computations}
Our algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers (DF) and the post-dominance frontiers (PDF) of the CFG. These points,
being structural properties of a graph, do not change during code-motion. A
post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it is the basic block on which the basic
block containing I is control dependent.

@@

\begin{verbatim}
struct CHIArg {
  // Edge destination (shows the direction of flow),
  // may not be where the I is.
  BasicBlock *Dest;
  // The instruction CHI-arg refers to.
  Instruction *I;
  // The global value number of I.
  GVNType VN;
};
\end{verbatim}

@@

\begin{verbatim}
// Compute barriers for both hoistable and sinkable
// candidates.
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

    // Find the last barrier below which instructions
    // can be sunk. If there was no barrier in B, any
    // instruction satisfying other legality checks
    // can be sunk.
    if barrier_found:
      for each instruction I in B in reverse order:
        if I does not guarantee progress:
          mark I as a sink barrier
          break
\end{verbatim}

@@

\begin{verbatim}
B0: b = m
    c = n
    if p is true then goto B1 else goto B2
B1: a0 = b<kill> + c<kill>
    goto B3
B2: a1 = b<kill> + c<kill>
    goto B3
B3: ...
\end{verbatim}

@@

\begin{verbatim}
B0: i0 = load B
B1: i1 = load A
    c1 = i1 + 10
    d1 = i0 + 20
    goto B3
B2: i2 = load A
    c2 = i2 + 10
    d2 = i0 + 20
    goto B3

B3: PHIc1, c2)
    PHI(d1, d2)
\end{verbatim}

@@

\subsection{Finding candidates to schedule}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers.

The current implementation of \GVN{} in \LLVM{} has some limitations when it
comes to loads and stores so we compute the \GVN{} of loads and stores
separately.  Our solution is to value number the address from where the value of
a load is to be read from memory. For stores, we value number the address as
well as the value to be stored at that address. Another limitation of the
current \GVN{} implementation in \LLVM{} is that the instructions dependent on
the loads will not get numbered correctly, and so after hoisting all candidates
we need to rerun the \GVN{} analysis in order to discover new candidates now
available after having hoisted load instructions.

The process of computing \GVN{} can be on-demand, as we come across an
instruction, or precomputed, computing \GVN{} of all the instructions
beforehand. Which process to choose is determined by the scope of code-hoisting
we want to perform. In a pessimistic approach, we want to hoist a limited set of
instructions from the sibling branches as we iterate the DFS tree bottom-up, it
is sufficient to compute \GVN{} values on-demand. Whereas, in the optimistic
approach, as described in Section~\ref{subsec:optimistic}, we want to move as
many instructions as possible, and it would require \GVN{} values to be
precomputed.

\subsection{Code generation for the optimistic \gcm{} algorithm}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting or sinking. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed, and the \SSA{} form is restored by
updating the intermediate representation (IR) to reflect the changes. At the
same time \MemorySSA{} is updated as well.

After one iteration of algorithm runs through the entire function, it creates
more opportunities for \emph{higher ranked} computations
\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
pass, and so we rerun the code-hoisting algorithm until there are no more
instructions left to be hoisted.  Obviously, this is not the most optimal
approach and can be improved by ranking the computations \cite{rosen1988global},
or by improving the \GVN{} analysis to correctly populate congruence classes as
the program is modified by the code generation.

Finally after the transformation is done, we verify a set of post-conditions to
establish that program invariants are maintained: e.g., consistency of use-defs,
and \SSA{} semantics.

The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm as mentioned earlier in
Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
\gcm{} of congruent instructions which uses the liveness analysis as
illustrated by \citet{das2012} and ranking expressions explained by
\citet{rosen1988global}.

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining.) The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic-block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

\begin{verbatim}
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CFG)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
   // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  doCodeSinking()
\end{verbatim}

We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm \emph{doGCM}
prefers hoisting (\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).

\begin{verbatim}
int doCodeHoisting(VNs)
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

@@

  for each CHI-nodes in the CFG:
    Remove the ones with missing entries
    Remove CHI-arg if it fails legality checks

  for each CHI-nodes in the CFG:
    if CHI  does not have missing entries:
      Perform code hoisting

  return number of sinked instructions
\end{verbatim}

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted) which are encountered
subsequently as the algorithm visits instructions in the increasing order of
their ranks.

%\newpage
After no more candidates are hoistable, sinking is performed. Sinking is only
performed once because currently there are very few sinkable candidates
(Figure~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
\GCM{}) has a code-sinking which already sinks several instructions.

For sinking we reuse the $\phi$ nodes instead of inserting another expression nodes
like in \cite{ssapre} because $\phi$ nodes ensure availability of definitions, and
hence downward safety, as long as all instructions corresponding to a $\phi$ are
congruent. Also, the $\phi$ nodes represent a complete set of instructions which
could be potentially sunk for code-size optimizations because $\phi$ nodes
represent all values flowing out from a basic block into its dominance-frontier.

\begin{verbatim}
int doCodeSinking()
  // Find sinkable instructions
  For each value number VN > 1 instructions:
    if there are two instructions with same immediate
        successor(S) as post-dominator:
      if both are only used in the same $\phi$-node of S
          && dependencies of both are available at S:
         mark instructions as sinkable

  // Sink the instructions and update statistics
  For each pair (I1, I2) of sinkable instructions:
    Move I1 to the sink point
    // the sink point is just after all
    // the PHI-nodes in the post-dominator
    update all the uses of PHI node (PN) with I1
    Remove I2 and PN
    if I1 is a memory reference:
      update MemorySSA of I1
      update MemorySSA of removed instructions
        to point to the MemorySSA reference of I1
      remove MemorySSA reference of all
        others which were deleted
    update statistics

  return number of sinked instructions
\end{verbatim}

>>>>

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{graphviz}
\usepackage{auto-pst-pdf}
\usepackage{etoolbox}
\usepackage{flushend}
\usepackage{needspace}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
  %columns=fixed,
  %fontadjust=true,
  basewidth=0.4em,
  language=C,
  mathescape
}
\usepackage[final]{pdfpages}

@@

\begin{lstlisting}
float fun(float d, float min, float max, float a) {
  float tmin, tmax;
  float inv = 1.0f / d;
  if (inv >= 0) {
    tmin = (min - a) * inv;
    tmax = (max - a) * inv;
  } else {
    tmin = (max - a) * inv;
    tmax = (min - a) * inv;
  }
  return tmax + tmin;
}
\end{lstlisting}

@@

Now if the computation of $tmax$ and $tmin$ are hoisted outside the
conditionals, the C code version would look like:
\begin{lstlisting}
float fun(float d, float min, float max, float a) {
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  float tmin = x, tmax = y;
  if (inv < 0) { tmin = y; tmax = x; }
  return tmax + tmin;
}
\end{lstlisting}

@@

\section{Factoring reachability of expressions}
Our algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers (DF) and the post-dominance frontiers (PDF) of the CFG. These points,
being structural properties of a graph, do not change during code-motion. A
post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it is the basic block on which the basic
block containing I is control dependent.

@@

\begin{lstlisting}
struct CHIArg {
  // Edge destination (shows the direction of flow),
  // may not be where the I is.
  BasicBlock *Dest;
  // The instruction CHI-arg refers to.
  Instruction *I;
  // The global value number of I.
  GVNType VN;
};
\end{lstlisting}

@@

\begin{lstlisting}
// Compute barriers for both hoistable and sinkable
// candidates.
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

    // Find the last barrier below which instructions
    // can be sunk. If there was no barrier in B, any
    // instruction satisfying other legality checks
    // can be sunk.
    if barrier_found:
      for each instruction I in B in reverse order:
        if I does not guarantee progress:
          mark I as a sink barrier
          break
\end{lstlisting}

@@

\begin{lstlisting}
B0: b = m
    c = n
    if p is true then goto B1 else goto B2
B1: a0 = b<kill> + c<kill>
    goto B3
B2: a1 = b<kill> + c<kill>
    goto B3
B3: ...
\end{lstlisting}

@@

\begin{lstlisting}
B0: i0 = load B
B1: i1 = load A
    c1 = i1 + 10
    d1 = i0 + 20
    goto B3
B2: i2 = load A
    c2 = i2 + 10
    d2 = i0 + 20
    goto B3

B3: PHIc1, c2)
    PHI(d1, d2)
\end{lstlisting}

@@

\section{A fast algorithm for code motion of congruent computations}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for code motion. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed.

\begin{lstlisting}
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CFG)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
    // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  // Code-sinking only once
  doCodeSinking()
\end{lstlisting}

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). If the dependency of a hoistable candidate is in the
same basic-block as the candidate, then the dependency must also be hoistable,
otherwise hoisting will be illegal or would require a complicated code
generation to make it legal. The current algorithm discards cases if the
dependency is neither hoistable nor rematerializable. We collect the \GVN{} of
all the instructions in the function and iterate on the list of instructions
having identical \GVN{}s. The algorithm \emph{doGCM} prefers hoisting
(\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).

\begin{lstlisting}
int doCodeHoisting(VNs)
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

@@

  for each CHI-nodes in the CFG:
    for each CHI-arg in CHI-node:
      Remove CHI-arg if it fails legality checks

  for each CHI-nodes in the CFG:
    if CHI  does not have missing entries:
      Perform code hoisting
      update data structures and statistics

  return number of hoisted instructions
\end{lstlisting}

Code hoisting opens new opportunities for other higher ranked
\cite{rosen1988global} (depended on candidates which got hoisted) which are
encountered subsequently as the algorithm visits instructions in the increasing
order of their ranks. Some of the congruent higher ranked computations do not
get equal value numbers so we rerun the code-hoisting algorithm until there are
no more instructions left to be hoisted. The hoisting converges quickly with 2-3
iterations.

\begin{lstlisting}
int doCodeSinking()
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

  for each VN in VNs
    for each instruction with the same VN
      Find its dominance-frontier (DF)
      Insert the PHI-node at DF

  for each PHI-nodes in the CFG:
    for each PHI-arg in PHI-node:
      Remove PHI-arg if it fails legality checks

  for each PHI-nodes in the CFG:
    if PHI does not have missing entries:
      Perform code sinking
      if a $\phi$ has same entries, reassign
      update data structures and statistics

      return number of sinked instructions
\end{lstlisting}

After no more candidates are hoistable, sinking is performed. It is only
performed once because it may not open new opportunities for sinkable
candidates. For sinking we iterate on the $\Phi$ nodes. After each code-motion
the \SSA{} form is restored by updating the intermediate representation (IR) to
reflect the changes. At the same time \MemorySSA{} is updated as well. Both the
data structures can be updated very efficiently. After one full iteration of
hoisting, the \GVN{} needs to be recomputed to explore new hoistable candidates.
Finally after the transformation is done, we verify a set of post-conditions to
establish that program invariants are maintained: e.g., consistency of use-defs,
and \SSA{} semantics.
